Distributional Trees and Forests

### 2017-11-27 pkg/disttree/inst/CMStatistics2017/slides.Rnw
slides for CMStatistics 2017 in London
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1759 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2da9a09935c92c5cfb76ed4efb7b17250e26aeb1

  dist_list_normal <- list(family.name = "Normal Distribution",
                           ddist = ddist, 
                           sdist = sdist, 
                           hdist = hdist,
                           pdist = pdist,
                           qdist = qdist,
                           rdist = rdist,
                           link = link, 
                           linkfun = linkfun, 
                           linkinv = linkinv, 
                           linkinvdr = linkinvdr,
                           startfun = startfun,
                           mle = mle
  )
}

@@

  dist_list_cens_normal <- list(family.name = "censored Normal Distribution",
                                ddist = ddist, 
                                sdist = sdist, 
                                hdist = hdist,
                                pdist = pdist,
                                qdist = qdist,
                                rdist = rdist,
                                link = link, 
                                linkfun = linkfun, 
                                linkinv = linkinv, 
                                linkinvdr = linkinvdr,
                                startfun = startfun,
                                mle = mle
  )
}

>>>>

  dist_list_normal <- list(family.name = "Normal Distribution",
                           ddist = ddist, 
                           sdist = sdist, 
                           hdist = hdist,
                           pdist = pdist,
                           qdist = qdist,
                           rdist = rdist,
                           link = link, 
                           linkfun = linkfun, 
                           linkinv = linkinv, 
                           linkinvdr = linkinvdr,
                           startfun = startfun,
                           mle = mle,
                           gamlssobj = FALSE,
                           censored = FALSE
  )
}

@@

  dist_list_cens_normal <- list(family.name = "censored Normal Distribution",
                                ddist = ddist, 
                                sdist = sdist, 
                                hdist = hdist,
                                pdist = pdist,
                                qdist = qdist,
                                rdist = rdist,
                                link = link, 
                                linkfun = linkfun, 
                                linkinv = linkinv, 
                                linkinvdr = linkinvdr,
                                startfun = startfun,
                                mle = mle,
                                gamlssobj = FALSE,
                                censored = TRUE
  )
}

### 2017-11-27 pkg/disttree/inst/CMStatistics2017/slides_rain.Rnw
slides for CMStatistics 2017 in London
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1759 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2da9a09935c92c5cfb76ed4efb7b17250e26aeb1

## HCL palette
pal <- hcl(c(10, 128, 260), 100, 50)
names(pal) <- c("forest", "tree", "gamlss")

pallight <- hcl(c(10, 128, 260), 100, 50, alpha = 0.25)
names(pallight) <- c("forest", "tree", "gamlss")

@@

@@

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<plotsim_RMSE_par, fig=TRUE, echo=FALSE>>=
plot_rmse(simlist, type = "par")
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Simulation}
\begin{center}
\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
<<plotsim_RMSE_exp, fig=TRUE, echo=FALSE>>=
plot_rmse(simlist, type = "exp")
@
\end{center}
\end{frame}

>>>>

## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
names(pal) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")

pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
names(pallight) <- c("forest", "tree", "gamlss", "gamboostLSS", "EMOS")

@@

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Precipitation}
\vspace*{0.2cm}
Data:
\begin{itemize}
\item Daily observations from 95 stations: \\
24h precipitation sums from July of 28 years.
\item Numerical ensemble predictions of meteorological quantities\\
\vspace{0.1cm}
$\rightarrow$ Covariates:\\
\begin{itemize}
\item Total precipitation
\item Convective available potential energy
\item Downwards short wave radiation flux (``sunshine'')
\item Mean sea level pressure 
\item Mean sea level pressure difference
\item Preciptable water
\item 2m maximum temperature
\item Total column-integrated condensate
\item Temperature 
\item Temperature differences in altitude
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Precipitation}
\vspace*{0.2cm}
Application for one station:
\begin{itemize}
\item Learn forest model on data from 24 years (1985--2008).
\item Evaluate on 4 years (2009--2012).
\end{itemize}
\end{frame}

<<pred_axams, eval=TRUE, echo=FALSE, results=hide>>=
#setwd("~/svn/partykit/pkg/disttree/inst/CMStatistics2017")
load("data_Axams.rda")

# tree and forest formula
{
  dt.formula <- df.formula <- 
      robs ~ tppow_mean + tppow_sprd + tppow_min + tppow_max + 
      tppow_mean0612 + tppow_mean1218 + tppow_mean1824 + tppow_mean2430 + 
      tppow_sprd0612 + tppow_sprd1218 + tppow_sprd1824 + tppow_sprd2430 + 
      capepow_mean + capepow_sprd + capepow_min + capepow_max + 
      capepow_mean0612 + capepow_mean1218 + capepow_mean1224 + capepow_mean1230 +
      capepow_sprd0612 + capepow_sprd1218 + capepow_sprd1224 + capepow_sprd1230 +
      dswrf_mean_mean + dswrf_mean_max +  
      dswrf_sprd_mean + dswrf_sprd_max + 
      msl_mean_mean + msl_mean_min + msl_mean_max + 
      msl_sprd_mean + msl_sprd_min + msl_sprd_max +
      pwat_mean_mean + pwat_mean_min + pwat_mean_max + 
      pwat_sprd_mean + pwat_sprd_min + pwat_sprd_max +
      tmax_mean_mean + tmax_mean_min + tmax_mean_max +
      tmax_sprd_mean + tmax_sprd_min + tmax_sprd_max +
      tcolc_mean_mean + tcolc_mean_min + tcolc_mean_max +
      tcolc_sprd_mean + tcolc_sprd_min + tcolc_sprd_max +
      t500_mean_mean + t500_mean_min + t500_mean_max +
      t700_mean_mean + t700_mean_min + t700_mean_max +
      t850_mean_mean + t850_mean_min + t850_mean_max +
      t500_sprd_mean + t500_sprd_min + t500_sprd_max +
      t700_sprd_mean + t700_sprd_min + t700_sprd_max +
      t850_sprd_mean + t850_sprd_min + t850_sprd_max +
      tdiff500850_mean + tdiff500850_min + tdiff500850_max +
      tdiff700850_mean + tdiff700850_min + tdiff700850_max +
      tdiff500700_mean + tdiff500700_min + tdiff500700_max +
      msl_diff

learndata <- data_axams[data_axams$year < 109,]
testdata <- data_axams[data_axams$year %in% c(109, 110, 111, 112),]

##############################################################
# fitting the models
set.seed(7)

df <- distforest(df.formula, data = learndata, family = dist_list_cens_normal, 
                 type.tree = "ctree", ntree = 100, censtype = "left", censpoint = 0,
                 control = ctree_control(teststat = "quad", testtype = "Univ",
                                         intersplit = TRUE,
                                         mincriterion = 0, minsplit = 50,
                                         minbucket = 20), mtry = 27)

#### prepare data for plot of estimated density functions
# predictions for one day (in each of the four years) 
# (19th of July 2011 is missing)
pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),

pdays <- if(pday<19) c(pday, pday + 31, pday + 62, pday + 92) else c(pday, pday + 31, pday + 61, pday + 92)
pdf <- predict(df, newdata = testdata[pdays,], type = "parameter")
df_mu <- pdf$mu
df_sigma <- pdf$sigma
df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
cbind(df_exp, testdata[pdays,"robs"])

# plot predicted distributions together with observations
#set.seed(res$call$seedconst)
set.seed(7)
x <- c(0.01, sort(runif(500,0.01,8)))
y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
y4 <- crch::dcnorm(x, mean = df_mu[4], sd = df_sigma[4], left = 0)
dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})

# point mass (slightly shifted)
pm1 <- c(0.05, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
pm2 <- c(0.01, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
pm3 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))
pm4 <- c(-0.07, crch::dcnorm(-1, mean = df_mu[4], sd = df_sigma[4], left = 0))

# predictions
pred1 <- c(testdata[pdays,"robs"][1], crch::dcnorm(testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
pred2 <- c(testdata[pdays,"robs"][2], crch::dcnorm(testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
pred3 <- c(testdata[pdays,"robs"][3], crch::dcnorm(testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))
pred4 <- c(testdata[pdays,"robs"][4], crch::dcnorm(testdata[pdays,"robs"][4], mean = df_mu[4], sd = df_sigma[4], left = 0))

#legendheight
lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
lh4 <- crch::dcnorm(0.01, mean = df_mu[4], sd = df_sigma[4], left = 0)
@

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Precipitation}
\begin{center}
\vspace*{-0.2cm}
<<plot_pred_axams_24July, fig=TRUE, echo=FALSE, eval=TRUE, height=5.2>>=
plot(x = x, y = y1, type = "l", col = "red", 
     main = paste0("Axams, July ", pday), ylab = "Density", 
     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
     #xlab = expression(Total~precipitation~group("[", mm^(1/1.6)~per~day,"]")),
     #{\frac{1}{1.6}}$,
     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
     xlim = c(-1.5,8))

lines(x = x, y = y2, type = "l", col = "blue")
lines(x = x, y = y3, type = "l", col = "darkgreen")
lines(x = x, y = y4, type = "l", col = "purple")
legend('topright', c("predicted distribution", "point mass at censoring point", "observation"),
       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)

# plot point mass
lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)

points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)

# plot predictions
points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)

lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)

# add labels
text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Precipitation}
\vspace*{0.2cm}
Application for one station:
\begin{itemize}
\item Learn forest model on data from 24 years.
\item Evaluate on 4 years.
\item 10 times 7-fold cross validation.
\item Compare to commonly used methods:\\
\begin{itemize}
\item Prespecified GAM
\item Boosted GAM
\item Ensemble Model Output Statistics (EMOS) 
\end{itemize}
\end{itemize}
\end{frame}

<<echo=FALSE, results=hide>>=
#### cross validation rain
if(file.exists("rain_Axams_7x10.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_7x10.rda")
  load("rain_Axams_7x10.rda")
} else {
  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_cross_7x10.R")
  source("rain_cross_7x10.R")
  library("gamlss.cens")
  gen.cens("NO", type = "left")
  rainres <- rain_cross(stationname = "Axams", 
                        seedconst = 7, ntree = 100,
                        forest_mtry = 27,
                        gamboost_cvr = TRUE,
                        frac = FALSE)

  save(rainres, file = "rain_Axams_7x10.rda")
}

rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres)-1)

for(i in 1:(length(rainres)-1)){
  rain_crps[i,] <- colMeans(rainres[[i]]$crps, na.rm = TRUE)
}

colnames(rain_crps) <- colnames(rainres[[1]]$rmse)

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Precipitation}
\begin{center}
\vspace*{-0.2cm}
<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=5.2>>=
boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
        main = "Cross Validation (with reference model EMOS)",
        names = c("Forest", "prespecified GAM", "boosted GAM"),
        ylab = "CRPS skills score", col = "lightgray") 
abline(h = 0, col = pal["EMOS"], lwd = 2)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Precipitation}
\vspace*{0.2cm}
Application for all stations:
\begin{itemize}
\item Learn forest model on data from 24 years (1985--2008).
\item Evaluate on 4 years (2009--2012).
\item Compare to commonly used methods:\\
\begin{itemize}
\item Prespecified GAM
\item Boosted GAM
\item Ensemble Model Output Statistics (EMOS) 
\end{itemize}
\end{itemize}
\end{frame}

<<echo = FALSE>>=
#### prediction over all stations 24 - 4
if(file.exists("rain_pred_24to4.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
  load("rain_pred_24to4.rda")
} else {

  source("rain_pred_24to4.R")
  library("gamlss.cens")
  gen.cens("NO", type = "left")

  res <- rain_pred(seedconst = 7,
                   gamboost_cvr = TRUE,
                   frac = FALSE)

  save(res, file = "rain_pred_24to4.rda")
}

crps_all <- res[[1]]$results["crps",]
for(i in 2:(length(res)-1))  crps_all <- rbind(crps_all, res[[i]]$results["crps",])

colnames(crps_all) <- colnames(res[[1]]$results)

# skills score
s <- 1 - crps_all[, 2:4]/crps_all[,6]
colnames(s) <- c("Forest", "prespecified GAM", "boosted GAM")

## prepare data for map which shows where distforest performed better than gamlss or gamboostLSS based on the crps

crps <- crps_all[,c("distforest", "gamlss", "gamboostLSS", "EMOS log")]

## best method
bst <- apply(crps, 1, which.min)

## distance of forest to best other method
dst <- crps[,1] - crps[cbind(1:nrow(crps), apply(crps[, -1], 1, which.min) + 1)]

## breaks/groups
brk <- c(-0.1, -0.05, -0.005, 0.005, 0.05, 0.1)
#brk <- c(-0.1, -0.05, -0.01, 0.01, 0.05, 0.1)
grp <- cut(dst, breaks = brk)

## HCL colors (relatively flashy, essentially CARTO Tropic)
clr <- colorspace::diverge_hcl(5, h = c(195, 325), c = 80, l = c(50, 90), power = 1.3)

library("raster") # dem (digital elevation model)
library("sp")     # gadm www.gadm.org/country

load("../draft/plot_map_rain/demo/tirol.gadm.rda")
load("../draft/plot_map_rain/demo/tirol.dem.rda")
load("../draft/plot_map_rain/demo/ehyd.statlist.rda")

# Create SpatialPointsDataFrame from station list
sp <- SpatialPointsDataFrame(subset(ehyd.statlist[res$complete_stations,],
                                    select=c(lon,lat)),
                             data = subset(ehyd.statlist[res$complete_stations,],
                                           select = -c(lon,lat)),
                             proj4string = crs(tirol.dem))

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Precipitation}
\begin{center}
\vspace*{0cm}

<<map, fig=TRUE, echo=FALSE, width=10, height = 6.5>>=

layout(cbind(1, 2), width = c(9, 1))
par(mar = c(5,4,4,0.1))
raster::image(tirol.dem, col = rev(gray.colors(100)),
              main="Observation Stations", ylab = "Latitude", xlab = "Longitude", 
              xlim = c(9.8,13.2), 
              ylim = c(46.6, 47.8), 
              cex.lab=1.5, cex.axis=1.2, cex.main=2)
plot(tirol.gadm, add = TRUE)
points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.9)
legend("topleft", pch=c(21,24,25,22), legend = c("Forest", "prespecified GAM", "boosted GAM", "EMOS"), cex = 1.2, bty = "n")

par(mar = c(0.5,0.2,0.5,2.3))
## legend
plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
     xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
axis(4, at = brk, las = 1, mgp=c(0,0,-1.2), cex = 1.4)

@
\end{center}
\end{frame}

@@
